D3064R0: OOTA Execution is Provably Vacuous

### 2023-12-06 ootavacuous.tex
Add verbiage and tools for dependency classification
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

We could follow this long-established tradition of proposing
solutions to the OOTA problem without a definition of this problem.
However, we (perhaps foolishly) felt the need to provide a clear
definition,

We therefore define an OOTA cycle (also known as a causal cycle) as a
cycle in the memory-model sense involving only reads-from links, that
is, the value stored in on thread being read by the next thread.
Within each thread in the cycle, the value stored depends on
the value loaded.
This dependency might be a control dependency
(see Listing~\ref{lst:OOTA Cycle})
or it might be a data dependency
(see Listing~\ref{lst:Causality Test Case 4} on
page~\pageref{lst:Causality Test Case 4}).\footnote{
	Sufficiently motivated readers should have no problem constructing
	examples involving address dependencies.}
@@@ There must be at least two distinct values such that if a given load
returns a given value, that value forces the cycle to execute so as to
cause the corresponding store's value to be equal to that given value, and
for that store to occur early enough for the load to return that value.

>>>>

One aspect of this complexity is the fact that semantic dependencies
are a function not strictly of the source code, but of a particular
execution.
Consider for example:

\begin{quote}
	\co{x = y * z;}
\end{quote}

There is a semantic dependency from \co{y} to \co{x} if and only if the
value of \co{z} is non-zero.
Otherwise, in the absence of observable behavior (such as volatile
accesses), the implementation is within its rights to discard this
statement and act as if later accesses to \co{x} were instead the
constant zero, at least up to the next assignment to \co{x}.

Similarly:

\begin{quote}
	\co{x = y + z;}
\end{quote}

There is a semantic dependency from both \co{y} and \co{z} if and only
if these values do not cancel.
However, if the implementation knows that the value of \co{y} is always
the negative of \co{x}, there is no semantic dependency and the
implementation can again act as if later accesses to \co{x} were
instead the constant zero.

And perhaps worse:

\begin{quote}
	\co{x = horribly_complex(y, z);}
\end{quote}

Here, the presence or absence of a semantic dependency from \co{y} to
\co{x} depends not only on the value of \co{z}, but also the code in
a horribly complex function.

Fortunately, there are now tools that can help.
The SAT-solver-based \co{cbmc} tool, which is readily available,\footnote{
	On Debian-based distributions, via \co{sudo apt install cbmc}.}
has been used to mechanically verify signicant portions of Linux-kernel
RCU from the C-language source
code~\cite{LihaoLiang2016VerifyTreeRCU,LanceRoy2017CBMC-SRCU}.
Another tool, Nidhugg, which is based on partial-order
reduction, has been used to carry out a similar
verification~\cite{MichalisKokologiannakis2017NidhuggRCU,SMC-TreeRCU,MichalisKokologiannakis2019RCUstatelessModelCheck}.
Both tools can easily check whether or not a given dependency is semantic.
% @@@ More on cbmc in an appendix.

However, as will be seen in later sections, in some special but
commonly occurring circumstances, there are easier ways.

% @@@ We could follow this long-established tradition of proposing
% solutions to the OOTA problem without a definition of this problem.
% However, we (perhaps foolishly) felt the need to provide a clear
% definition,

% We therefore define an OOTA cycle (also known as a causal cycle) as a
% cycle in the memory-model sense involving only reads-from links, that
% is, the value stored in on thread being read by the next thread.
% Within each thread in the cycle, the value stored depends on
% the value loaded.
% This dependency might be a control dependency
% (see Listing~\ref{lst:OOTA Cycle})
% or it might be a data dependency
% (see Listing~\ref{lst:Causality Test Case 4} on
% page~\pageref{lst:Causality Test Case 4}).\footnote{
%	Sufficiently motivated readers should have no problem constructing
%	examples involving address dependencies.}
% @@@ There must be at least two distinct values such that if a given load
% returns a given value, that value forces the cycle to execute so as to
% cause the corresponding store's value to be equal to that given value, and
% for that store to occur early enough for the load to return that value.
